---
title: "data_science_viz"
author: "Laura Sørine Voldgaard"
date: "2026-05-24"
output: html_document
---

## Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
getwd()
```

```{r}
# packages
pacman::p_load(tidyverse)

```


## Data prep

```{r}
# load data

# for MSE plot
MSE_results_linear_lasso <- read_csv("results/cv_df.csv") # MSE results for linear and lasso models
MSE_results_bart <- read_csv("results/BART_ucloud/cv_df.csv") # MSE results for bart

# for prediction plot
null_pred <- read_csv("results/pred_uncertainty/null_pred.csv") # predictions for 2016, 2020, and 2024 election for the null model
#linear_pred <- read_csv("results/pred_uncertainty/pred_df.csv") # predictions for 2016, 2020, and 2024 election for some linear models
lasso_pred <- read_csv("results/pred_uncertainty/boot_df_lasso_pers_all_int.csv") # predictions for 2016, 2020, and 2024 election for some lasso models
bart_pred <- read_csv("results/pred_uncertainty/bart_pred_df.csv") # predictions for 2016, 2020, and 2024 election for some bart models

# for variable inclusion plot
variable_inclusion_bart <- read_csv("results/BART_ucloud/var_df.csv") # predictions for 2016, 2020, and 2024 election for some models


```



## Plot


### MSE results

```{r}
# combine plots
MSE_results_all <- bind_rows(MSE_results_linear_lasso, MSE_results_bart)
MSE_results_all
```




```{r}
# filter for the models we want to show
MSE_filtered <- MSE_results_all %>%
  filter(method %in% c("linear_null", 
                       "linear_all", 
                       "lasso_all_no_int", 
                       "lasso_all_all_int", 
                       "lasso_econ_PC2", 
                       "bart_all", 
                       "bart_PC3_econ", 
                       "bart_well",
                       "lasso_pers_all_int",
                       "lasso_econ_PC2"))


# plot 
MSE_results_v1 <- ggplot(MSE_filtered, aes(x = MSE, y = reorder(method, -MSE))) +
  geom_col(fill = "darkcyan", width = 0.6) +
  geom_text(aes(label = round(MSE, 2)), hjust = -0.2, size = 3.5) +
  labs(
    title = "Cross-Validation MSE by Model",
    x = "MSE",
    y = NULL
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(hjust = 0.5),
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text.y = element_text(color = "black")
  )

MSE_results_v1

```

null✔
linear_all✔

lasso_all✔
lasso_econ (best predicting) - lasso_econ_PC2✔
lasso_wellbeing (best predicting) - lasso_pers_all_int✔

bart_all✔
bart_econ (best predicting) - bart_PC3_econ✔
bart_wellbeing (best predicting) - bart_well✔







### Prediction results

```{r}
# prep null data
names(null_pred)

# only keep last 3 rows
null_pred_filtered <- tail(null_pred, 3)

# add year column
null_pred_filtered <- tail(null_pred, 3) %>%
  mutate(year = c(2016, 2020, 2024)) %>%
  select(year, everything())

null_pred_filtered # done

```


```{r}
# prep linear data

```



```{r}
# prep lasso data
lasso_pred

# remove columns
lasso_pred_filtered <- lasso_pred %>% select(-`Boot std`, -`Contains actual`)

# rename column
lasso_pred_filtered <- lasso_pred_filtered %>% 
  rename(actual = Actual,
         lasso_pers_all_int = Predicted,
         lasso_pers_all_int_lower = Lower,
         lasso_pers_all_int_upper = Upper)

lasso_pred_filtered

```



```{r}
# prep bart data
names(bart_pred)

bart_pred_filtered <- bart_pred %>%
  select(year, actual, bart_well, bart_well_lower, bart_well_upper, 
         bart_PC3_econ, bart_PC3_econ_lower, bart_PC3_econ_upper)

bart_pred_filtered # done
```


```{r}
# merge dataframes
all_preds <- null_pred_filtered %>%
  left_join(bart_pred_filtered, by = "year") %>% 
  left_join(lasso_pred_filtered, by = "year")


# remove extra "actual" columns
all_preds <- all_preds %>% select(-`actual.x`, -`actual.y`)

all_preds
```






```{r}

# define which columns are point predictions
# (everything except year, actual, and the lower/upper columns)
model_cols <- c("null", "bart_well", "bart_PC3_econ", "lasso_pers_all_int")

# pivot to long format with lower and upper attached to each model
pred_long <- model_cols %>%
  map_dfr(function(m) {
    all_preds %>%
      select(year, actual,
             predicted = all_of(m),
             lower     = all_of(paste0(m, "_lower")),
             upper     = all_of(paste0(m, "_upper"))) %>%
      mutate(model = m)
  })

# plot
predictions_v2 <- ggplot(pred_long, aes(x = year)) +
  # uncertainty ribbon for each model
  geom_ribbon(aes(ymin = lower, ymax = upper, fill = model), alpha = 0.2) +
  scale_y_continuous(limits = c(40, 60), oob = scales::squish) +
  # model prediction lines
  geom_line(aes(y = predicted, color = model), linetype = "dashed", linewidth = 1) +
  geom_point(aes(y = predicted, color = model), size = 4) +
  # actual values in black on top
  geom_line(aes(y = actual), color = "black", linewidth = 1.5) +
  geom_point(aes(y = actual), color = "black", size = 4) +
  labs(x = "Election Year", y = "Vote Share for Incumbent (%)", 
       color = "Model", fill = "Model") +
  theme_minimal(base_size = 13) +
  scale_x_continuous(breaks = c(2016, 2020, 2024))

predictions_v2
```

actual✔
null✔
lasso_econ_PC2✔
bart_wellbeing (personal)✔
bart_econ_PC2✔












### Variable inclusion plot

```{r}
# for bart_all
variable_inclusion_v1 <- variable_inclusion_bart %>%
  filter(!is.na(bart_all_inclusion)) %>%
  ggplot(aes(x = bart_all_inclusion, y = reorder(predictor, bart_all_inclusion))) +
  geom_col(fill = "darkcyan", width = 0.6) +
  geom_text(aes(label = round(bart_all_inclusion, 2)), hjust = -0.2, size = 3.5) +
  scale_x_continuous(limits = c(0, 40), oob = scales::squish) +
  labs(
    title = "Variable Inclusion Proportions (BART)",
    x = "Inclusion Proportion",
    y = NULL
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(hjust = 0.5),
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text.y = element_text(color = "black")
  )

variable_inclusion_v1

```




## Save plots

```{r}
# MSE results
ggsave("results/figures/MSE_results_v1.jpg", MSE_results_v1, width = 12, height = 10, dpi = 300)

# Predictions
ggsave("results/figures/predictions_v2.jpg", predictions_v1, width = 12, height = 10, dpi = 300)

# Variable inclusion
ggsave("results/figures/variable_inclusion_bart_v1.jpg", variable_inclusion_v1, width = 12, height = 10, dpi = 300)

```







### NYT plot
```{r}
# merge dataframes
all_preds <- null_pred_filtered %>%
  left_join(bart_pred_filtered, by = "year") %>% 
  left_join(lasso_pred_filtered, by = "year")


# remove extra "actual" columns
all_preds <- all_preds %>% select(-`actual.x`, -`actual.y`)

all_preds
```



"null", "bart_well", "bart_PC3_econ", "actual"

```{r}
all_preds_nyt <- all_preds %>%
  select(year, actual, 
         null, null_lower, null_upper,
         bart_well, bart_well_lower, bart_well_upper,
         bart_PC3_econ, bart_PC3_econ_lower, bart_PC3_econ_upper)

colnames(all_preds_nyt)
```


With uncertainty
```{r}
pred_long_simple <- bind_rows(
  all_preds_nyt %>% transmute(year, model = "Actual",           predicted = actual,        lower = NA,                 upper = NA),
  all_preds_nyt %>% transmute(year, model = "Null model",       predicted = null,          lower = null_lower,         upper = null_upper),
  all_preds_nyt %>% transmute(year, model = "BART (wellbeing)", predicted = bart_well,     lower = bart_well_lower,    upper = bart_well_upper),
  all_preds_nyt %>% transmute(year, model = "BART (economic)",  predicted = bart_PC3_econ, lower = bart_PC3_econ_lower, upper = bart_PC3_econ_upper)
) %>%
  mutate(model = factor(model, levels = c("Actual", "Null model",
                                          "BART (wellbeing)", "BART (economic)")))
```





```{r}
# THE ONE
predictions_bar <- ggplot(pred_long_simple,
                          aes(x = factor(year), y = predicted, fill = model)) +
  geom_bar(stat = "identity",
           position = position_dodge(width = 0.8),
           width = 0.7) +
  geom_errorbar(aes(ymin = lower, ymax = upper),
                position = position_dodge(width = 0.8),
                width = 0.2,
                color = "gray30",
                linewidth = 0.7) +
  geom_hline(yintercept = 50,
             linetype = "dashed",
             color = "gray40") +
  scale_y_continuous(limits = c(0, 65),
                     breaks = seq(0, 65, 10),
                     expand = c(0, 0)) +
  scale_fill_manual(values = c(
    "Actual"           = "black",
    "Null model"       = "#b0b0b0",
    "BART (wellbeing)" = "#707070",
    "BART (economic)"  = "#d8d8d8"
  ),
  labels = c(
    "Actual"           = "True vote share",
    "Null model"       = "Model with mean vote share accross 12 elections",
    "BART (wellbeing)" = "Machine Learning model based on wellbeing factors",
    "BART (economic)"  = "Machine Learning model based on macroeconomic factors"
    
  )) +
  labs(x = NULL,
       y = "Vote share for Incumbent (%)",
       fill = NULL,
       title = "How well did our models predict incumbent vote share?",
       subtitle = "Predicted vs. actual popular vote share for the incumbent party: Presidential elections 2016, 2020, and 2024",
       caption = "Error bars reflect uncertainty intervals in each model's prediction, showing the range of plausible predicted values") +
  theme_minimal(base_size = 13) +
  theme(
    legend.position    = "top",
    legend.direction   = "horizontal",
    panel.grid.major.x = element_blank(),
    panel.grid.minor   = element_blank(),
    plot.title         = element_text(face = "bold", size = 20, hjust = 0),
    plot.subtitle      = element_text(color = "gray40", size = 14, hjust = 0),
    plot.caption       = element_text(color = "gray50", size = 11, hjust = 0),
    axis.text.x = element_text(face = "bold", size = 13, margin = margin(t = 8))  )

predictions_bar
```

---------------------------------------------------------------------------------------



DID NOT USE
```{r}
#install.packages("cowplot")
#install.packages("magick")
library(cowplot)
library(magick)
library(grid)

# load images directly - they already have circles
donkey_grob   <- rasterGrob(as.raster(image_read("donkey.png")),   interpolate = TRUE)
elephant_grob <- rasterGrob(as.raster(image_read("elephant.png")), interpolate = TRUE)

# build plot with blank x-axis text
predictions_bar <- ggplot(pred_long_simple,
                          aes(x = factor(year), y = predicted, fill = model)) +
  geom_bar(stat = "identity",
           position = position_dodge(width = 0.8),
           width = 0.7) +
  geom_errorbar(aes(ymin = lower, ymax = upper),
                position = position_dodge(width = 0.8),
                width = 0.2,
                color = "gray30",
                linewidth = 0.7) +
  geom_hline(yintercept = 50,
             linetype = "dashed",
             color = "gray40") +
  scale_y_continuous(limits = c(0, 65),
                     breaks = seq(0, 65, 10),
                     expand = c(0, 0)) +
  scale_fill_manual(values = c(
    "Actual"           = "black",
    "Null model"       = "#999999",
    "BART (wellbeing)" = "#D4A017",
    "BART (economic)"  = "#4A4A4A"
  ),
  labels = c(
    "Actual"           = "True vote share",
    "Null model"       = "Model with mean vote share across 12 elections",
    "BART (wellbeing)" = "Machine Learning model based on wellbeing factors",
    "BART (economic)"  = "Machine Learning model based on macroeconomic factors"
  )) +
  labs(x = NULL,
       y = "Vote share for Incumbent (%)",
       fill = NULL,
       title = "How well did our models predict incumbent vote share?",
       subtitle = "Predicted vs. actual popular vote share for the incumbent party, 2016-2024",
       caption = "Error bars reflect uncertainty intervals in each model's prediction, showing the range of plausible predicted values") +
  theme_minimal(base_size = 13) +
  theme(
    legend.position    = "top",
    legend.direction   = "horizontal",
    legend.text        = element_text(size = 10),
    panel.grid.major.x = element_blank(),
    panel.grid.minor   = element_blank(),
    plot.title         = element_text(face = "bold", size = 14, hjust = 0),
    plot.subtitle      = element_text(color = "gray40", size = 11, hjust = 0),
    plot.caption       = element_text(color = "gray50", size = 9, hjust = 0),
    axis.text.x        = element_blank(),
    plot.margin        = margin(t = 5, r = 5, b = 40, l = 5)
  )

# combine plot with images and year labels below
final_plot <- ggdraw(predictions_bar) +
  draw_grob(donkey_grob,   x = 0.18, y = -0.06, width = 0.12, height = 0.12) +
  draw_grob(elephant_grob, x = 0.47, y = -0.06, width = 0.12, height = 0.12) +
  draw_grob(donkey_grob,   x = 0.76, y = -0.06, width = 0.12, height = 0.12) +
  draw_label("2016", x = 0.24, y = 0.04, size = 11, fontface = "bold") +
  draw_label("2020", x = 0.53, y = 0.04, size = 11, fontface = "bold") +
  draw_label("2024", x = 0.82, y = 0.04, size = 11, fontface = "bold") +
  draw_label("Error bars reflect uncertainty intervals in each model's prediction,\nshowing the range of plausible predicted values",
             x = 0.98, y = 0.84, size = 8, color = "gray50",
             hjust = 1, vjust = 1)

final_plot

# save with bigger dimensions to give room for circles
#ggsave("predictions_nyt.png", final_plot, width = 12, height = 8, dpi = 300)
```


